---
title: "Bulk RNA-seq Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bulk RNA-seq Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

This tutorial shows the two bulk RNA-seq configurations covered here in one
continuous R workflow:

1. quantify a sample from short reads;
2. add transcript-level long-read counts and quantify the same sample again.

The second configuration reuses the index and short-read counts created for the
first, making it easy to compare the two results.

## What This Tutorial Produces

By the end, you will have:

- an `mpaqt_index` object for the reference transcriptome;
- an `mpaqt_counts_sr` object prepared from paired short-read FASTQ files;
- an `mpaqt_counts_lr` object prepared from a Bambu transcript-count table;
- `sr_result`, based on short reads only; and
- `sr_lr_result`, based on short and long reads together.

The workflow is:

```text
reference + short reads  ->  sr_result
                     + long-read counts  ->  sr_lr_result
```

## Required Inputs

For the short-read configuration, you need:

- a GTF annotation and matching transcriptome FASTA;
- paired short-read FASTQ files; and
- `kallisto` and `bustools` available on your system.

For the short-plus-long-read configuration, you also need a transcript-count
table produced by Bambu. Its first column contains transcript IDs, and its
second column contains counts for the sample used here.

Transcript IDs must use the same convention in the GTF, transcriptome FASTA,
and Bambu table, including version suffixes when present.

## Create or Load the Index

Load MPAQT and create the index once for a reference:

```{r bulk-create-index}
library(mpaqt)

index <- mpaqt_index(
    annotation = "reference/gencode.annotation.gtf",
    transcriptome = "reference/gencode.transcripts.fa",
    output_file = "reference/mpaqt.index.rds",
    read_length = 91L,
    store_gtf = TRUE,
    threads = 8L
)
```

Set `read_length` to the length of the transcript sequence represented by one
short-read mate. `store_gtf = TRUE` lets the same index support the optional
raw-FLNC route shown later. Reuse the saved index for later samples:

```{r bulk-load-index}
index <- mpaqt_read_index("reference/mpaqt.index.rds")
```

The [index reference](../reference/mpaqt_index.html) describes additional ways
to create an index, including starting from a genome or an existing kallisto
index.

## Prepare Short Reads

Prepare equivalence-class counts from paired FASTQ files:

```{r bulk-prepare-short}
sr_counts <- mpaqt_prepare_short_reads(
    index = index,
    fastq_1 = "data/sample_R1.fastq.gz",
    fastq_2 = "data/sample_R2.fastq.gz",
    output_dir = "results/short-read",
    threads = 8L
)
```

This step runs kallisto and bustools, returns an `mpaqt_counts_sr` object, and
saves `results/short-read/mpaqt.short_read.rds` for reuse.

## Quantify with Short Reads

Run the short-read-only configuration:

```{r bulk-quant-short}
sr_result <- mpaqt_quant(
    index = index,
    sr_counts = sr_counts
)
```

`sr_result` is an `mpaqt_quant_result` object containing transcript-level TPM
values, abundances, convergence information, and fitted values.

```{r bulk-view-short}
head(tpm(sr_result))
```

## Prepare Long Reads

For the second configuration, prepare long-read counts from a Bambu table such
as:

```text
transcript_id,sample1
ENST00000456328.2,18
ENST00000450305.2,6
ENST00000488147.1,0
```

Pass that table to `mpaqt_prepare_long_reads()`:

```{r bulk-prepare-long}
lr_counts <- mpaqt_prepare_long_reads(
    index = index,
    bambu_counts = "data/bambu_counts.csv",
    output_dir = "results/long-read"
)
```

MPAQT matches the table's transcript IDs to the index order and saves
`results/long-read/mpaqt.long_read.rds`. Transcripts absent from the table
receive zero counts; transcript IDs that are not in the index are ignored, and
substantial mismatches trigger a warning.

## Quantify with Short and Long Reads

Reuse `index` and `sr_counts`, then add `lr_counts`:

```{r bulk-quant-short-long}
sr_lr_result <- mpaqt_quant(
    index = index,
    sr_counts = sr_counts,
    lr_counts = lr_counts
)
```

You now have both configurations in separate objects: `sr_result` for short
reads alone and `sr_lr_result` for short and long reads together.

## Inspect and Save Results

Use the result accessors to inspect the second configuration:

```{r bulk-inspect-results}
head(tpm(sr_lr_result))
head(abundances(sr_lr_result))
head(gene_abundances(sr_lr_result, index))
```

`gene_abundances()` needs the same index used for quantification so it can map
transcripts to genes.

Save both results with distinct prefixes:

```{r bulk-save-results}
mpaqt_save_result(
    sr_result,
    output_dir = "results/quantification",
    prefix = "sample.short"
)

mpaqt_save_result(
    sr_lr_result,
    output_dir = "results/quantification",
    prefix = "sample.short_long"
)
```

Each call writes a full `.quant.rds` result and a `.quant.csv` table of TPM
values. The files can be loaded later with `readRDS()` and
`data.table::fread()`, respectively.

## Alternative Input Formats

The examples above use the most direct path from paired FASTQ files and a
Bambu table. If those inputs have already been processed, reuse the existing
files instead.

Load prepared short-read counts from RDS:

```{r bulk-short-rds}
sr_counts <- mpaqt_prepare_short_reads(
    index = index,
    rds_file = "results/short-read/mpaqt.short_read.rds"
)
```

Or prepare them from matching BUS and equivalence-class files:

```{r bulk-short-bus}
sr_counts <- mpaqt_prepare_short_reads(
    index = index,
    bus_file = "data/output.bus",
    ec_file = "data/matrix.ec",
    output_dir = "results/short-read"
)
```

Load prepared long-read counts from RDS:

```{r bulk-long-rds}
lr_counts <- mpaqt_prepare_long_reads(
    index = index,
    rds_file = "results/long-read/mpaqt.long_read.rds"
)
```

Raw FLNC reads can also be aligned and quantified before MPAQT prepares the
long-read count object:

```{r bulk-long-flnc}
lr_counts <- mpaqt_prepare_long_reads(
    index = index,
    flnc_fastq = "data/sample.flnc.fastq.gz",
    genome = "reference/genome.fa",
    output_dir = "results/long-read",
    threads = 8L
)
```

This route additionally requires minimap2, samtools, and the Bioconductor
package Bambu. See the
[short-read](../reference/mpaqt_prepare_short_reads.html) and
[long-read](../reference/mpaqt_prepare_long_reads.html) references for input
details.

## Troubleshooting

**Transcript IDs do not match.** Confirm that the GTF, transcriptome FASTA, and
count table come from the same reference release and either all retain or all
remove transcript version suffixes.

**kallisto or bustools is not found.** Check the external tools described in
the [installation article](installation.html), then restart R if your `PATH`
changed.

**The Bambu table has multiple samples.** The preparation function uses the
second column. Supply a table containing the transcript-ID column and the one
sample column you want to quantify.

**A saved object cannot be loaded.** Make sure the RDS contains the matching
MPAQT object type and was prepared against the same index.

## Next Steps

- See the [single-cell workflow](single-cell-workflow.html) for cluster-level
  quantification and the additional single-cell configurations.
- Read [quantification methods](quantification-methods.html) for the statistical
  model and algorithm details.
- Use the [function reference](../reference/index.html) for complete argument
  descriptions.
