---
title: "Bulk RNA-seq Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bulk RNA-seq Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article provides comprehensive documentation of the MPAQT bulk RNA-seq workflow, including all input pathways and algorithm parameters.

## Workflow Overview

```
BULK WORKFLOW STAGES
====================

┌─────────────────────────────────────────────────────────────────────────────┐
│                           STAGE 1: INDEX CREATION                           │
│                                                                             │
│   Input: GTF annotation + transcriptome (FASTA, genome, or pre-built)       │
│   Output: mpaqt_index object with P matrices, distances, covariates         │
│                                                                             │
│   mpaqt_index() ─────► mpaqt.index.rds                                      │
└─────────────────────────────────────────────────────────────────────────────┘
                                      │
                                      ▼
┌─────────────────────────────────────────────────────────────────────────────┐
│                        STAGE 2: SHORT-READ PROCESSING                       │
│                                                                             │
│   Input: FASTQ files, BUS/EC files, or pre-computed RDS                     │
│   Output: Equivalence class counts (mpaqt_counts_sr)                        │
│                                                                             │
│   mpaqt_prepare_short_reads() ─────► mpaqt.short_read.rds                   │
└─────────────────────────────────────────────────────────────────────────────┘
                                      │
                                      ▼
┌─────────────────────────────────────────────────────────────────────────────┐
│                       STAGE 3: LONG-READ PROCESSING                         │
│                             (OPTIONAL)                                      │
│                                                                             │
│   Input: FLNC FASTQ, Bambu counts, or pre-computed RDS                      │
│   Output: Transcript counts (mpaqt_counts_lr)                               │
│                                                                             │
│   mpaqt_prepare_long_reads() ─────► mpaqt.long_read.rds                     │
└─────────────────────────────────────────────────────────────────────────────┘
                                      │
                                      ▼
┌─────────────────────────────────────────────────────────────────────────────┐
│                          STAGE 4: QUANTIFICATION                            │
│                                                                             │
│   Input: Index + short-read counts + long-read counts (optional)            │
│   Output: Transcript abundances, TPM, uncertainty                           │
│                                                                             │
│   mpaqt_quant() ─────► mpaqt.quant.rds + mpaqt.quant.csv                    │
└─────────────────────────────────────────────────────────────────────────────┘
```

---

## Input Pathways Summary

MPAQT supports multiple input pathways for flexibility. Choose the pathway that matches your available data:

### Index Creation Pathways

| Pathway | Inputs | External Tools | Bioconductor Packages |
|---------|--------|----------------|----------------------|
| **A: Standard** | GTF + transcriptome FASTA | kallisto | Biostrings, rtracklayer |
| **B: Genome** | GTF + genome (BSgenome) | kallisto | Biostrings, rtracklayer, BSgenome, GenomicRanges |
| **C: Pre-built** | GTF + FASTA + kallisto.idx | none | rtracklayer |

### Short-Read Processing Pathways

| Pathway | Inputs | External Tools |
|---------|--------|----------------|
| **A: FASTQ** | R1.fastq.gz + R2.fastq.gz | kallisto, bustools |
| **B: BUS** | output.bus + matrix.ec | bustools |
| **C: RDS** | mpaqt.short_read.rds | none |

### Long-Read Processing Pathways

| Pathway | Inputs | External Tools | Bioconductor Packages |
|---------|--------|----------------|----------------------|
| **A: FLNC** | flnc.fastq.gz + genome | minimap2, samtools | bambu, BSgenome |
| **B: Bambu counts** | counts_transcript.csv | none | none |
| **C: RDS** | mpaqt.long_read.rds | none | none |

> **Recommended for long-read integration:** Use **Pathway B (Bambu counts)**. Run Bambu externally and provide the counts CSV. This requires no additional tools or Bioconductor packages.

---

## Progressive Workflow Examples

This section demonstrates MPAQT capabilities through progressively complex examples. Each level shows **CLI first, then R API**.

### Level 1: Short-Read Only

The simplest MPAQT workflow uses only short-read data.

#### Using CLI

```bash
# Step 1: Create index (one-time setup)
mpaqt index \
  --transcriptome reference/transcripts.fa \
  --annotation reference/annotation.gtf \
  --output my_index.rds \
  --threads 8

# Step 2: Process short reads
mpaqt prepare-sr my_index.rds \
  --fastq-r1 sample_R1.fastq.gz \
  --fastq-r2 sample_R2.fastq.gz \
  --output-dir results/ \
  --threads 8

# Step 3: Quantify
mpaqt quant my_index.rds results/mpaqt.sr.rds \
  --output-dir results/
```

#### Using R API

```{r level1-r, eval=FALSE}
library(mpaqt)

# Step 1: Create index
index <- mpaqt_index(
  annotation = "reference/annotation.gtf",
  transcriptome = "reference/transcripts.fa",
  output_file = "my_index.rds",
  threads = 8
)

# Step 2: Process short reads
sr_counts <- mpaqt_prepare_short_reads(
  index = index,
  fastq_1 = "sample_R1.fastq.gz",
  fastq_2 = "sample_R2.fastq.gz",
  output_dir = "results/",
  threads = 8
)

# Step 3: Quantify
result <- mpaqt_quant(index, sr_counts)

# Extract TPM
tpm_values <- tpm(result)
head(tpm_values)
```

---

### Level 2: Short-Read + Long-Read Integration

Adding long-read data improves quantification accuracy, especially for genes with complex splicing patterns. MPAQT uses long-read counts as prior information in a MAP estimation framework.

#### Using CLI

```bash
# Prepare long-read counts (from Bambu output)
mpaqt prepare-lr my_index.rds \
  --counts-file bambu_counts.csv \
  --output-dir results/

# Quantify with both data types
mpaqt quant my_index.rds results/mpaqt.sr.rds \
  --lr-counts results/mpaqt.lr.rds \
  --output-dir results/
```

#### Using R API

```{r level2-r, eval=FALSE}
# Prepare long-read counts (from Bambu)
lr_counts <- mpaqt_prepare_long_reads(
  index = index,
  counts_file = "bambu_counts.csv",
  output_dir = "results/"
)

# Quantify with both data types
result <- mpaqt_quant(
  index = index,
  sr_counts = sr_counts,
  lr_counts = lr_counts
)

# Results now incorporate long-read information
tpm_values <- tpm(result)
```

---

### Level 3: Adding Positional Bias Correction

For 3' poly(A)-selected libraries, reads are enriched near transcript 3' ends. Enable bias correction for more accurate abundance estimates.

#### Using CLI

```bash
mpaqt quant my_index.rds results/mpaqt.sr.rds \
  --lr-counts results/mpaqt.lr.rds \
  --bias 3p \
  --bins 50 \
  --output-dir results/
```

#### Using R API

```{r level3-r, eval=FALSE}
result <- mpaqt_quant(
  index = index,
  sr_counts = sr_counts,
  lr_counts = lr_counts,
  positional_bias = "3p",
  n_bins = 50
)
```

> **Note:** Use `positional_bias = "5p"` for 5'-biased protocols (e.g., some single-cell methods).

---

### Level 4: Full Configuration with Prior Integration

For publication-quality results, enable all features including uncertainty estimation. Prior integration is automatic when long-read data is provided.

#### Using CLI

```bash
mpaqt quant my_index.rds results/mpaqt.sr.rds \
  --lr-counts results/mpaqt.lr.rds \
  --bias 3p \
  --bins 50 \
  --prior-start 25 \
  --compute-uncertainty \
  --output-dir results/
```

#### Using R API

```{r level4-r, eval=FALSE}
result <- mpaqt_quant(
  index = index,
  sr_counts = sr_counts,
  lr_counts = lr_counts,
  positional_bias = "3p",
  n_bins = 50,
  prior_model = "shrinkage",
  prior_start = 25,
  compute_uncertainty = TRUE
)

# Access all results
tpm_values <- tpm(result)
se <- uncertainty(result)
gene_tpm <- gene_abundances(result, index)

# Save results
mpaqt_save_result(result, "results/")
```

> **Note:** The `--prior-start` option controls when the EM algorithm starts incorporating the prior (default: iteration 25). Prior integration is automatic when long-read data is provided.

---

## Detailed Reference

The following sections provide comprehensive documentation for each stage.

---

## Stage 1: Index Creation

The MPAQT index contains all reference data needed for quantification:

- **Kallisto index**: For short-read pseudoalignment
- **P matrices**: Mapping equivalence classes to transcripts with positional information
- **Distance matrices**: 5' and 3' distances for bias correction
- **Covariate matrix**: Transcript length, GC content, biotype
- **Transcript-gene mappings**: For gene-level aggregation

### Input Pathways

```
INDEX CREATION PATHWAYS
=======================

Pathway A: GTF + Transcriptome FASTA (Recommended)
──────────────────────────────────────────────────
  mpaqt_index(
      annotation   = "gencode.v44.gtf",
      transcriptome = "gencode.v44.transcripts.fa"
  )

Pathway B: GTF + BSgenome (Extract transcriptome)
─────────────────────────────────────────────────
  mpaqt_index(
      annotation = "gencode.v44.gtf",
      genome     = "BSgenome.Hsapiens.UCSC.hg38"
  )

Pathway C: GTF + FASTA + Pre-built Kallisto Index
─────────────────────────────────────────────────
  mpaqt_index(
      annotation     = "gencode.v44.gtf",
      transcriptome  = "gencode.v44.transcripts.fa",
      kallisto_index = "prebuilt.kallisto.idx"
  )
```

### Function Reference: `mpaqt_index()`

```r
mpaqt_index(
  annotation,           # Path to GTF annotation file (required)
  transcriptome = NULL, # Path to transcriptome FASTA
  genome = NULL,        # BSgenome object or package name
  kallisto_index = NULL,# Path to pre-built Kallisto index
  output_file = "mpaqt.index.rds",
  read_length = 75L,    # Expected read length
  stride = 1L,          # Stride for read simulation
  threads = 1L,
  chunk_size = 10000000L,
  store_sequences = TRUE,  # Store sequences for long-read workflows
  store_gtf = TRUE,        # Store GTF for long-read workflows
  temp_dir = tempdir(),
  keep_temp = FALSE,
  verbose = TRUE
)
```

### Key Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `annotation` | (required) | GTF file with transcript and gene annotations |
| `transcriptome` | NULL | FASTA file with transcript sequences |
| `genome` | NULL | BSgenome name for transcript extraction |
| `kallisto_index` | NULL | Pre-built Kallisto index to reuse |
| `read_length` | 75 | Expected read length (affects P matrix construction) |
| `store_sequences` | TRUE | Bundle transcriptome for long-read workflows |
| `store_gtf` | TRUE | Bundle GTF for long-read workflows |

### Example: Complete Index Creation

```r
library(mpaqt)

# Create index from Gencode reference
index <- mpaqt_index(
  annotation = "reference/gencode.v44.annotation.gtf",
  transcriptome = "reference/gencode.v44.transcripts.fa",
  output_file = "my_project/mpaqt.index.rds",
  read_length = 75,
  threads = 8,
  store_sequences = TRUE,  # Enable for long-read workflows
  store_gtf = TRUE
)

# View index summary
print(index)
# mpaqt_index object
#   Transcripts: 253,863
#   Genes: 63,234
#   Equivalence classes: 1,847,293
#   Has kallisto index: yes (bundled)
#   Has transcriptome: yes
#   Has GTF: yes
```

### Index Internals

The index creation process:

1. **Parse GTF**: Extract transcript-gene mappings and biotypes
2. **Process FASTA**: Compute GC content and append polyA tails
3. **Build Kallisto index**: For pseudoalignment
4. **Simulate reads**: Generate reads from every position in each transcript
5. **Run kallisto bus**: Determine equivalence class memberships
6. **Build P matrices**: Store EC-transcript counts with positional distances
7. **Bundle**: Create single portable file with all components

---

## Stage 2: Short-Read Processing

Convert Illumina short-read data to equivalence class counts.

### Input Pathways

```
SHORT-READ PROCESSING PATHWAYS
==============================

Pathway A: Raw FASTQ Files (Runs kallisto bus internally)
─────────────────────────────────────────────────────────
  mpaqt_prepare_short_reads(
      index   = index,
      fastq_1 = "sample_R1.fastq.gz",
      fastq_2 = "sample_R2.fastq.gz"
  )

Pathway B: Pre-computed BUS + EC Files
──────────────────────────────────────
  mpaqt_prepare_short_reads(
      index    = index,
      bus_file = "kallisto_out/output.bus",
      ec_file  = "kallisto_out/matrix.ec"
  )

Pathway C: Pre-computed RDS File
────────────────────────────────
  mpaqt_prepare_short_reads(
      index    = index,
      rds_file = "previous_run/mpaqt.short_read.rds"
  )
```

### Function Reference: `mpaqt_prepare_short_reads()`

```r
mpaqt_prepare_short_reads(
  index,              # mpaqt_index object
  fastq_1 = NULL,     # Forward FASTQ (R1)
  fastq_2 = NULL,     # Reverse FASTQ (R2)
  bus_file = NULL,    # Pre-computed BUS file
  ec_file = NULL,     # Pre-computed EC mapping file
  rds_file = NULL,    # Pre-computed RDS file
  output_dir = tempdir(),
  threads = 1L,
  verbose = TRUE
)
```

### Key Parameters

| Parameter | Description |
|-----------|-------------|
| `fastq_1`, `fastq_2` | Paired-end FASTQ files (gzipped OK) |
| `bus_file`, `ec_file` | Pre-computed kallisto bus output |
| `rds_file` | Previously saved counts (fastest pathway) |
| `threads` | Parallel threads for kallisto |

### Example: Process Short Reads

```r
# From raw FASTQ files
sr_counts <- mpaqt_prepare_short_reads(
  index = index,
  fastq_1 = "data/sample_R1.fastq.gz",
  fastq_2 = "data/sample_R2.fastq.gz",
  output_dir = "results/",
  threads = 8
)

# View counts summary
print(sr_counts)
# mpaqt_counts_sr object
#   Equivalence classes: 847,293
#   Total counts: 45,238,193
#   Sample ID: sample

# Counts are automatically saved to output_dir
# Load later with:
sr_counts <- mpaqt_read_short_read_counts("results/mpaqt.short_read.rds")
```

### Output Structure

The `mpaqt_counts_sr` object contains:

- `ec_counts`: Named numeric vector of EC counts
- `total_reads`: Total number of reads processed
- `sample_id`: Sample identifier
- `input_files`: List of input file paths
- `processing_info`: Metadata about processing

---

## Stage 3: Long-Read Processing (Optional)
Long-read data provides direct transcript evidence to resolve ambiguities.

> **Easiest approach:** If you have pre-computed Bambu counts (e.g., from running Bambu externally), use the **Bambu counts pathway**. This requires no additional tools or Bioconductor packages.

### Input Pathways

```
LONG-READ PROCESSING PATHWAYS
=============================

Pathway A: FLNC FASTQ (Full pipeline: minimap2 + Bambu)
───────────────────────────────────────────────────────
  mpaqt_prepare_long_reads(
      index      = index,
      flnc_fastq = "isoseq/sample.flnc.fastq",
      genome     = "BSgenome.Hsapiens.UCSC.hg38"
  )

Pathway B: Pre-computed Bambu Counts
────────────────────────────────────
  mpaqt_prepare_long_reads(
      index        = index,
      bambu_counts = "bambu_output/counts_transcript.csv"
  )

Pathway C: Pre-computed RDS File
────────────────────────────────
  mpaqt_prepare_long_reads(
      index    = index,
      rds_file = "previous_run/mpaqt.long_read.rds"
  )
```

### Function Reference: `mpaqt_prepare_long_reads()`

```r
mpaqt_prepare_long_reads(
  index,                  # mpaqt_index object
  flnc_fastq = NULL,      # FLNC FASTQ from Iso-Seq
  genome = NULL,          # Genome for minimap2 alignment
  bambu_counts = NULL,    # Pre-computed Bambu counts
  rds_file = NULL,        # Pre-computed RDS file
  output_dir = tempdir(),
  threads = 1L,
  minimap_preset = "splice:hq",  # For Iso-Seq HiFi reads
  discovery = FALSE,             # Enable novel transcript discovery
  sample_id = NULL,
  verbose = TRUE
)
```

### Key Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `flnc_fastq` | NULL | Full-length non-chimeric reads from Iso-Seq |
| `genome` | NULL | Genome FASTA or BSgenome for alignment |
| `bambu_counts` | NULL | Pre-computed transcript counts from Bambu |
| `minimap_preset` | "splice:hq" | Preset for minimap2 (splice:hq for HiFi) |
| `discovery` | FALSE | Allow Bambu novel transcript discovery |

### Example: Process Long Reads

```r
# From FLNC FASTQ (full pipeline)
lr_counts <- mpaqt_prepare_long_reads(
  index = index,
  flnc_fastq = "data/sample.flnc.fastq.gz",
  genome = "BSgenome.Hsapiens.UCSC.hg38",
  output_dir = "results/",
  threads = 8
)

# From pre-computed Bambu output
lr_counts <- mpaqt_prepare_long_reads(
  index = index,
  bambu_counts = "bambu/counts_transcript.csv",
  output_dir = "results/"
)

# View counts summary
print(lr_counts)
# mpaqt_counts_lr object
#   Transcripts with counts: 45,293
#   Total counts: 2,134,567
#   Sample ID: sample

# Load later with:
lr_counts <- mpaqt_read_long_read_counts("results/mpaqt.long_read.rds")
```

---

## Stage 4: Quantification

Run the MPAQT EM algorithm to estimate transcript abundances.

### Algorithm Overview

MPAQT maximizes the following objective:

```
L(beta) = L_SR(beta) + L_LR(beta) - R(beta)

Where:
  L_SR: Short-read Poisson likelihood over equivalence classes
  L_LR: Long-read Poisson likelihood over transcripts
  R:    L2 regularization penalty
```

The EM algorithm iterates between:

1. **E-step**: Assign read probabilities to transcripts
2. **M-step**: Update abundance estimates

### Quantification Options

```
QUANTIFICATION CONFIGURATIONS
=============================

Basic: Short-reads only
───────────────────────
  mpaqt_quant(index, sr_counts)

With Long-reads:
────────────────
  mpaqt_quant(index, sr_counts, lr_counts)

With Positional Bias Correction:
────────────────────────────────
  mpaqt_quant(index, sr_counts, positional_bias = "3p")

With Shrinkage Prior:
─────────────────────
  mpaqt_quant(index, sr_counts, prior_model = "shrinkage")

Full Configuration:
───────────────────
  mpaqt_quant(
      index, sr_counts, lr_counts,
      positional_bias = "3p",
      prior_model = "shrinkage",
      compute_uncertainty = TRUE
  )
```

### Function Reference: `mpaqt_quant()`

```r
mpaqt_quant(
  index,                      # mpaqt_index object
  sr_counts,                  # mpaqt_counts_sr object
  lr_counts = NULL,           # mpaqt_counts_lr object (optional)
  positional_bias = NULL,     # "3p", "5p", or NULL
  n_bins = 50L,               # Distance bins for bias correction
  prior_model = NULL,         # "shrinkage", "long_read", or NULL
  normalize = "tpm",          # "tpm", "depth", or "none"
  max_iter = 100L,            # Maximum EM iterations
  tolerance = 1e-4,           # Convergence tolerance
  prior_start = 25L,          # Iteration to start using prior
  convergence_start = 25L,    # Iteration to check convergence
  compute_uncertainty = FALSE,# Compute standard errors
  verbose = TRUE
)
```

### Key Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `positional_bias` | NULL | Bias correction type: "3p" (poly-A), "5p", or NULL |
| `n_bins` | 50 | Number of distance bins for bias modeling |
| `prior_model` | NULL | Prior for regularization |
| `max_iter` | 100 | Maximum EM iterations |
| `tolerance` | 1e-4 | Log-likelihood change threshold for convergence |
| `compute_uncertainty` | FALSE | Bootstrap-based standard errors |

### Positional Bias Correction

For 3' poly(A) selected libraries, reads are enriched near the 3' end of transcripts. The `positional_bias = "3p"` option models this bias:

```r
# 3' bias correction (recommended for most RNA-seq)
result <- mpaqt_quant(
  index = index,
  sr_counts = sr_counts,
  positional_bias = "3p",
  n_bins = 50
)
```

### Example: Complete Quantification

```r
# Full quantification with all features
result <- mpaqt_quant(
  index = index,
  sr_counts = sr_counts,
  lr_counts = lr_counts,        # Optional long-read integration
  positional_bias = "3p",       # 3' bias correction
  prior_model = "shrinkage",    # Regularization
  max_iter = 100,
  tolerance = 1e-4,
  compute_uncertainty = TRUE,   # Bootstrap SEs
  verbose = TRUE
)

# View result summary
print(result)
# mpaqt_quant_result object
#   Transcripts: 253,863
#   Converged: TRUE (47 iterations)
#   Log-likelihood: -1,234,567.89
#   Has uncertainty: yes
```

---

## Extracting Results

### Result Accessors

```r
# Transcript-level TPM values
tpm_values <- tpm(result)
head(tpm_values)
# ENST00000456328.2 ENST00000450305.2 ENST00000488147.1
#          0.12345           2.34567           0.00000

# Raw abundances (beta values)
abundances <- abundances(result)

# Standard errors (if computed)
if (has_uncertainty(result)) {
  se <- uncertainty(result)
}

# Gene-level abundances
gene_tpm <- gene_abundances(result)
```

### Saving Results

```r
# Save full result object + CSV
mpaqt_save_result(
  result = result,
  output_dir = "results/",
  prefix = "sample1",
  include_abundances = TRUE,
  include_uncertainty = TRUE
)

# Creates:
#   results/sample1.quant.rds  (full R object)
#   results/sample1.quant.csv  (TPM values)
```

### Export Utilities

```r
# Export to CSV with custom columns
export_results_csv(
  result = result,
  output_file = "results/sample1_tpm.csv",
  include_abundances = FALSE,
  include_uncertainty = TRUE
)

# Combine multiple samples
results_list <- list(
  sample1 = result1,
  sample2 = result2,
  sample3 = result3
)

combined <- combine_quant_results(results_list)
# data.table with transcript_id, sample1, sample2, sample3 columns
```

---

## Complete Example Workflow

### Using CLI

```bash
#!/bin/bash
# Complete bulk RNA-seq MPAQT workflow

# Stage 1: Create index (run once per reference)
mpaqt index \
  --transcriptome reference/gencode.v44.transcripts.fa \
  --annotation reference/gencode.v44.annotation.gtf \
  --output reference/mpaqt.index.rds \
  --read-length 75 \
  --threads 8

# Stage 2: Process short reads
mpaqt prepare-sr reference/mpaqt.index.rds \
  --fastq-r1 data/sample_R1.fastq.gz \
  --fastq-r2 data/sample_R2.fastq.gz \
  --output-dir results/ \
  --threads 8

# Stage 3: Process long reads (optional)
mpaqt prepare-lr reference/mpaqt.index.rds \
  --counts-file data/bambu_counts.csv \
  --output-dir results/

# Stage 4: Quantify
mpaqt quant reference/mpaqt.index.rds results/mpaqt.sr.rds \
  --lr-counts results/mpaqt.lr.rds \
  --bias 3p \
  --compute-uncertainty \
  --output-dir results/

# Output files:
#   results/mpaqt.quant.rds   - Full result object
#   results/mpaqt.quant.csv   - TPM values
```

**Or use the one-command pipeline:**

```bash
mpaqt run reference/mpaqt.index.rds \
  --fastq-r1 data/sample_R1.fastq.gz \
  --fastq-r2 data/sample_R2.fastq.gz \
  --lr-counts-file data/bambu_counts.csv \
  --bias 3p \
  --output-dir results/ \
  --threads 8
```

### Using R API

```r
library(mpaqt)

#===============================================================================
# STAGE 1: Create Index (run once per reference)
#===============================================================================

index <- mpaqt_index(
  annotation = "reference/gencode.v44.annotation.gtf",
  transcriptome = "reference/gencode.v44.transcripts.fa",
  output_file = "reference/mpaqt.index.rds",
  read_length = 75,
  threads = 8,
  store_sequences = TRUE,
  store_gtf = TRUE
)

#===============================================================================
# STAGE 2: Process Short Reads
#===============================================================================

sr_counts <- mpaqt_prepare_short_reads(
  index = index,
  fastq_1 = "data/sample_R1.fastq.gz",
  fastq_2 = "data/sample_R2.fastq.gz",
  output_dir = "results/",
  threads = 8
)

#===============================================================================
# STAGE 3: Process Long Reads (Optional)
#===============================================================================

lr_counts <- mpaqt_prepare_long_reads(
  index = index,
  bambu_counts = "data/bambu_counts.csv",
  output_dir = "results/"
)

#===============================================================================
# STAGE 4: Quantify
#===============================================================================

result <- mpaqt_quant(
  index = index,
  sr_counts = sr_counts,
  lr_counts = lr_counts,
  positional_bias = "3p",
  compute_uncertainty = TRUE,
  verbose = TRUE
)

#===============================================================================
# STAGE 5: Extract and Save Results
#===============================================================================

# Get TPM values
tpm_values <- tpm(result)

# Get gene-level abundances
gene_tpm <- gene_abundances(result)

# Save everything
mpaqt_save_result(result, "results/", prefix = "sample")
```

---

## Advanced Topics

### Two-Phase Quantification

For more control, run pre-quantification and post-quantification separately:

```r
# Phase 1: Estimate positional bias weights
prequant <- mpaqt_prequant(
  index = index,
  sr_counts = sr_counts,
  positional_bias = "3p",
  n_bins = 50
)

# Save for reuse
mpaqt_save_prequant(prequant, "results/")

# Phase 2: Run main EM with pre-computed weights
result <- mpaqt_postquant(
  index = index,
  sr_counts = sr_counts,
  lr_counts = lr_counts,
  prequant = prequant,
  max_iter = 100
)
```

### Loading Saved Objects

```r
# Load index
index <- mpaqt_read_index("reference/mpaqt.index.rds")

# Load counts
sr_counts <- mpaqt_read_short_read_counts("results/mpaqt.short_read.rds")
lr_counts <- mpaqt_read_long_read_counts("results/mpaqt.long_read.rds")

# Load result
result <- mpaqt_read_result("results/sample.quant.rds")
```

### Multi-Sample Analysis

```r
# Process multiple samples
samples <- c("sample1", "sample2", "sample3")

results <- lapply(samples, function(s) {
  sr <- mpaqt_prepare_short_reads(
    index = index,
    fastq_1 = paste0("data/", s, "_R1.fastq.gz"),
    fastq_2 = paste0("data/", s, "_R2.fastq.gz"),
    output_dir = paste0("results/", s, "/")
  )

  mpaqt_quant(index, sr, positional_bias = "3p")
})
names(results) <- samples

# Combine into TPM matrix
tpm_matrix <- combine_quant_results(results)
```

---

## Troubleshooting

### Common Issues

**Error: Kallisto not found**
```
# Ensure kallisto is in PATH
Sys.which("kallisto")

# Or install via conda
# conda install -c bioconda kallisto
```

**Error: No matching ECs**
```
# Transcript IDs in FASTQ/counts must match GTF
# Check that transcript IDs are consistent
```

**Convergence issues**
```r
# Try increasing max_iter or adjusting tolerance
result <- mpaqt_quant(
  index, sr_counts,
  max_iter = 200,
  tolerance = 1e-3
)
```

### Performance Tips

1. **Use threads**: Set `threads` parameter for kallisto operations
2. **Reuse index**: Build index once, use for all samples
3. **Save intermediates**: Use `output_dir` to save counts for reuse
4. **Skip bias correction**: If not needed, omit `positional_bias`
