---
title: "Get Started with MPAQT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with MPAQT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

## Overview

**MPAQT** (Multi-Platform Aggregation and Quantification of Transcripts) is an R package for RNA-seq transcript quantification that integrates short-read and long-read sequencing data. By combining the depth of Illumina sequencing with the resolution of PacBio/ONT long reads, MPAQT achieves more accurate transcript-level abundance estimates than either platform alone.

### Key Features

- **Multi-platform integration**: Combine Illumina short-reads with PacBio/ONT long-reads in a unified probabilistic framework
- **Bulk and single-cell support**: Unified API for both analysis types
- **Positional bias correction**: Account for 3' or 5' sequencing biases common in poly(A) selection protocols
- **Uncertainty quantification**: Fisher information-based standard errors for abundance estimates
- **Flexible input options**: Start from FASTQ files, BUS files, or pre-computed counts
- **Command-line interface**: Shell-friendly workflow automation for pipeline integration

## Why MPAQT?

MPAQT achieves **state-of-the-art isoform-level quantification** by addressing fundamental limitations of existing methods.

### Generative Probabilistic Model

At the core of MPAQT is a generative model that connects **latent transcript abundances** to observed "observation units" (OUs). An observation unit is any entity we can directly quantify from RNA-seq reads, defined in a technology-dependent manner:

- **Long-read (LR) data**: Each transcript is one OU, resulting in a one-to-one relationship. The expected count scales linearly with abundance, with factors like transcript length and GC content affecting the slope.
- **Short-read (SR) data**: Equivalence classes (ECs) serve as OUs. Each transcript may connect to multiple OUs, and each OU may connect to multiple transcripts—MPAQT models expected counts as linear functions of all contributing transcript abundances.

Unlike alignment-free methods that discard multi-mapping information, MPAQT preserves the full probabilistic structure of read-transcript assignments.

### Equivalence Classes for Short Reads

Short reads are grouped into **equivalence classes (ECs)**—sets of reads that map to the same group of transcripts. This approach:

- Preserves multi-mapping information that other methods discard
- Enables efficient computation through sparse matrix operations
- Avoids false precision from incorrect read assignments

### Long-Read Integration via MAP Estimation

When long-read data is available, MPAQT uses **maximum a posteriori (MAP) estimation** with a log-normal prior informed by long-read counts:

- Direct transcript-level evidence from full-length reads
- Resolution of ambiguities that short reads alone cannot resolve
- Robust estimates even with moderate long-read coverage (~200K reads)
- Automatic learning of LR-specific biases (length, GC content)

### Validated Superior Performance

In benchmarks against kallisto, salmon, and RSEM:

- **13–32% more variance explained** in isoform-level abundance (simulations)
- **RT-qPCR validation**: Pearson r=0.81 for SR+LR vs r=0.33 for SR-only
- **Gene-level accuracy**: Pearson r=0.91 vs ground truth (MAQC data), with fewer outliers than competing tools
- Particularly beneficial for genes with **complex splicing patterns** (e.g., nervous system development genes with more exons and isoforms)

### Comprehensive Bias Correction

MPAQT accounts for multiple sources of bias:

- **Length bias**: Longer transcripts generate more reads
- **GC content bias**: Sequence-dependent amplification effects—longer and GC-poor transcripts are more likely captured by LR sequencing
- **3'/5' positional bias**: Read coverage variation along transcripts in poly(A)-selected libraries

## Installation

### Quick Install

```bash
# Step 1: Install R and compilation dependencies
conda install -c conda-forge r-base cmake icu zlib

# Step 2: Install required system tools
conda install -c bioconda kallisto bustools
```

```{r}
# Step 3: Install MPAQT R package
install.packages("pak")
pak::pak("csglab/mpaqt2")
```

### Alternative Installation Methods

MPAQT is also available via:

- **Docker**: `docker pull csglab/mpaqt:latest`
- **Apptainer**: `apptainer pull library://csglab/mpaqt:latest`
- **Conda**: `conda install -c csglab -c conda-forge -c bioconda r-mpaqt`

For detailed installation options including container variants (stable, full, dev), see the [Installation Guide](installation.html).

## Quick Start

### Minimal Example

```r
library(mpaqt)

# 1. Create index from reference files
index <- mpaqt_index(
  annotation = "reference/gencode.v44.annotation.gtf",
  transcriptome = "reference/gencode.v44.transcripts.fa",
  output_file = "my_index.mpaqt.rds"
)

# 2. Process short-read FASTQ files
sr_counts <- mpaqt_prepare_short_reads(
  index = index,
  fastq_1 = "reads/sample_R1.fastq.gz",
  fastq_2 = "reads/sample_R2.fastq.gz",
  output_dir = "output/"
)

# 3. Run quantification
result <- mpaqt_quant(index, sr_counts)

# 4. Extract results
tpm_values <- tpm(result)
head(tpm_values)
```

### With Long-Read Integration

```r
# Process long-read counts (from Bambu or similar)
lr_counts <- mpaqt_prepare_long_reads(
  index = index,
  counts_file = "longread/bambu_counts.csv",
  output_dir = "output/"
)

# Quantify with both data types
result <- mpaqt_quant(index, sr_counts, lr_counts)

# Results now incorporate long-read information
tpm_values <- tpm(result)
```

## Workflow Overview

MPAQT workflows consist of four main stages:

```
┌─────────────────┐
│  INDEX CREATION │  Create reference index from GTF + transcriptome
└────────┬────────┘
         │
         ▼
┌─────────────────┐
│  READ PROCESSING│  Convert FASTQ to equivalence class counts
└────────┬────────┘
         │
         ▼
┌─────────────────┐
│  QUANTIFICATION │  EM algorithm for abundance estimation
└────────┬────────┘
         │
         ▼
┌─────────────────┐
│  RESULTS        │  Extract TPM, abundances, uncertainty
└─────────────────┘
```

### CLI vs R API

MPAQT provides both a **command-line interface (CLI)** and an **R API**. All functionality is available through both interfaces:

| Task | CLI Command | R Function |
|------|-------------|------------|
| Create index | `mpaqt index` | `mpaqt_index()` |
| Process short reads (bulk) | `mpaqt prepare-sr` | `mpaqt_prepare_short_reads()` |
| Process short reads (SC) | `mpaqt prepare-sr --technology 10xv3` | `mpaqt_prepare_short_reads_sc()` |
| Process long reads | `mpaqt prepare-lr` | `mpaqt_prepare_long_reads()` |
| Quantify (bulk) | `mpaqt quant` | `mpaqt_quant()` |
| Quantify (single-cell) | `mpaqt quant-sc` | `mpaqt_quant_sc()` |
| Full pipeline | `mpaqt run` | (chain individual functions) |

**When to use CLI:**

- Integrating into shell pipelines (Snakemake, Nextflow)
- Batch processing multiple samples
- Running on HPC clusters

**When to use R API:**

- Interactive exploration
- Custom analysis workflows
- Integration with other R packages (Seurat, scanpy via reticulate)

### CLI Setup

The MPAQT CLI is installed with the R package at `inst/bin/mpaqt`.

```bash
# Find the installed location
Rscript -e "cat(system.file('bin', 'mpaqt', package = 'mpaqt'))"

# Add to PATH (add to ~/.bashrc for persistence)
export PATH="$PATH:$(Rscript -e 'cat(dirname(system.file(\"bin\", \"mpaqt\", package = \"mpaqt\")))')"

# Verify installation
mpaqt --version
mpaqt --help
```

### One-Command Pipeline with `mpaqt run`

For standard bulk RNA-seq workflows, `mpaqt run` combines all steps:

```bash
mpaqt run my_index.rds \
  --fastq-r1 sample_R1.fastq.gz \
  --fastq-r2 sample_R2.fastq.gz \
  --lr-counts-file bambu_counts.csv \
  --bias 3p \
  --output-dir results/ \
  --threads 8
```

This creates:
```
results/
├── mpaqt.sr.rds         # Short-read EC counts
├── mpaqt.lr.rds         # Long-read counts
├── mpaqt.quant.rds      # Full quantification result
└── mpaqt.quant.csv      # TPM values (for downstream tools)
```

Each stage offers multiple input pathways to accommodate different starting points. See the [Bulk Workflow](bulk-workflow.html) and [Single-Cell Workflow](single-cell-workflow.html) articles for detailed documentation, and the [CLI Reference](cli-reference.html) for complete command documentation.

## Core Concepts

### Equivalence Classes

MPAQT uses **equivalence classes** (ECs) for short-read quantification. An equivalence class represents a set of transcripts that a read could have originated from. This approach:

- Avoids expensive alignment to individual transcripts
- Preserves multi-mapping information
- Enables efficient EM algorithm computation

### The EM Algorithm

MPAQT uses an Expectation-Maximization algorithm to estimate transcript abundances:

**E-step**: Assign read probabilities to transcripts based on current abundance estimates

**M-step**: Update abundance estimates based on assigned reads

The algorithm iterates until convergence, incorporating:

- Short-read equivalence class counts
- Long-read direct transcript counts (optional)
- Positional bias correction (optional)
- L2 regularization to prevent overfitting

### Long-Read Integration

Long reads provide direct evidence of full-length transcripts, resolving ambiguities that short reads cannot. MPAQT integrates long-read counts as:

- **Prior information** for the EM algorithm
- **Direct counts** added to the likelihood function
- **Bias correction** factors from observed coverage patterns

## Function Reference

### Index Creation

| Function | Description |
|----------|-------------|
| `mpaqt_index()` | Create index from GTF + transcriptome |
| `mpaqt_read_index()` | Load a saved index |

### Short-Read Processing

| Function | Description |
|----------|-------------|
| `mpaqt_prepare_short_reads()` | Process bulk short-read FASTQ |
| `mpaqt_prepare_short_reads_sc()` | Process single-cell short-reads |
| `mpaqt_read_short_read_counts()` | Load saved short-read counts |

### Long-Read Processing

| Function | Description |
|----------|-------------|
| `mpaqt_prepare_long_reads()` | Process bulk long-read data |
| `mpaqt_prepare_long_reads_sc()` | Process single-cell long-reads |
| `mpaqt_read_long_read_counts()` | Load saved long-read counts |

### Quantification

| Function | Description |
|----------|-------------|
| `mpaqt_quant()` | Main quantification function |
| `mpaqt_quant_bulk()` | Alias for bulk quantification |
| `mpaqt_quant_sc()` | Single-cell quantification |
| `mpaqt_prequant()` | Pre-quantification (bias correction) |
| `mpaqt_postquant()` | Post-quantification (main EM) |

### Results

| Function | Description |
|----------|-------------|
| `abundances()` | Extract raw abundances |
| `tpm()` | Extract TPM values |
| `uncertainty()` | Extract standard errors |
| `gene_abundances()` | Aggregate to gene level |
| `mpaqt_save_result()` | Save results to disk |

## Next Steps

- **[Bulk Workflow](bulk-workflow.html)**: Detailed guide for bulk RNA-seq analysis
- **[Single-Cell Workflow](single-cell-workflow.html)**: Guide for single-cell analysis
- **[CLI Reference](cli-reference.html)**: Command-line interface documentation

## Developers

- Ali Saberi
- Michael Apostolides
- Jichen Wang
- Hamed S. Najafabadi

## Getting Help

If you encounter issues:

1. Check the function documentation: `?mpaqt_quant`
2. Review the workflow articles for your use case
3. Ensure external tools (kallisto, bustools) are properly installed
4. Report issues at: https://github.com/csglab/mpaqt2/issues
