---
title: "Algorithm and Mathematical Methods"
author: "MPAQT Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Algorithm and Mathematical Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

## Overview

MPAQT (Multi-Platform Aggregation and Quantification of Transcripts) integrates short-read and long-read RNA-seq data to improve transcript abundance estimation. This vignette provides a comprehensive mathematical description of the quantification algorithm.

### The Core Optimization Problem

MPAQT estimates transcript abundances $\boldsymbol{\beta} = (\beta_1, \ldots, \beta_T)$ by maximizing a composite objective function:

$$\max_{\boldsymbol{\beta}} L(\boldsymbol{\beta}) = L_{SR}(\boldsymbol{\beta}) + L_{LR}(\boldsymbol{\beta}) - R(\boldsymbol{\beta})$$

where:

- $L_{SR}(\boldsymbol{\beta})$: Short-read log-likelihood (equivalence class model)
- $L_{LR}(\boldsymbol{\beta})$: Long-read log-likelihood (direct transcript counts)
- $R(\boldsymbol{\beta})$: Regularization penalty (log-normal prior)


## Short-Read Likelihood ($L_{SR}$)

### Equivalence Classes

Short reads often map ambiguously to multiple transcripts due to shared sequences. Rather than discarding multi-mapping reads, MPAQT uses **equivalence classes** (ECs) - groups of reads that map to the same set of transcripts.

For each equivalence class $i$, we observe $n_i$ reads and have a probability matrix $\mathbf{P}$ where $P_{ij}$ is the probability that a read from transcript $j$ falls into EC $i$.

### The Multinomial Model

The expected read count for EC $i$ is:

$$y_i = \sum_{j=1}^{T} \beta_j \cdot P_{ij}$$

where $T$ is the total number of transcripts.

Under a Poisson approximation to the multinomial, the log-likelihood is:

$$L_{SR}(\boldsymbol{\beta}) = \sum_{i=1}^{E} \left[ n_i \log(y_i) - y_i \right]$$

where $E$ is the number of equivalence classes.

### Computing the P Matrix

The probability $P_{ij}$ that a fragment from transcript $j$ is assigned to EC $i$ depends on:

1. **Transcript length**: Longer transcripts generate more reads
2. **EC membership**: Which transcripts share the EC
3. **Effective length**: Adjusted for fragment size

```{r p-matrix-concept, echo=FALSE, eval=FALSE}
# Conceptual example of P matrix construction
# Rows = equivalence classes
# Columns = transcripts
# Values = probability of read from transcript mapping to EC

# Example with 3 transcripts (A, B, C) and 4 ECs
P_example <- matrix(
  c(
    1.0,   0,   0,    # EC1: unique to A
    0,   1.0,   0,    # EC2: unique to B
    0.5, 0.5,   0,    # EC3: shared A,B
    0.3, 0.3, 0.4     # EC4: shared A,B,C
  ),
  nrow = 4, byrow = TRUE,
  dimnames = list(
    paste0("EC", 1:4),
    c("Transcript_A", "Transcript_B", "Transcript_C")
  )
)
```


## Long-Read Likelihood ($L_{LR}$)

### Direct Transcript-Level Counts

Unlike short reads, long reads typically span full transcripts and can be unambiguously assigned. Let $c_j$ be the observed long-read count for transcript $j$.

### Coverage Probability Model

Long-read coverage varies across transcripts due to:

1. **3'/5' positional bias**: Degradation or priming effects
2. **Sequence-dependent effects**: GC content, secondary structure
3. **Technical variation**: Library prep, sequencing

We model the expected long-read count as:

$$\mathbb{E}[c_j] = \beta_j \cdot p_j \cdot d$$

where:

- $p_j$: Coverage probability for transcript $j$ (from GLM)
- $d$: Sequencing depth scale factor

### The Poisson Likelihood

The long-read log-likelihood is:

$$L_{LR}(\boldsymbol{\beta}) = \sum_{j=1}^{T} \left[ c_j \log(\beta_j \cdot p_j) - \beta_j \cdot p_j \right]$$

For transcripts without long-read coverage ($c_j = 0$), this term contributes:

$$-\beta_j \cdot p_j$$

which acts as a soft penalty discouraging high abundance estimates for uncovered transcripts.


## Regularization ($R$)

### The Log-Normal Prior

To prevent overfitting and stabilize estimates for lowly-expressed transcripts, MPAQT uses L2 regularization in log-space:

$$R(\boldsymbol{\beta}) = \frac{\lambda}{2} \sum_{j=1}^{T} \left( \log(\beta_j) - \mu_j \right)^2$$

This is equivalent to a log-normal prior:

$$\log(\beta_j) \sim \mathcal{N}(\mu_j, \sigma^2)$$

where $\sigma^2 = 1/\lambda$.

### Adaptive Prior Mean

The prior mean $\mu_j$ can be set:

1. **Global**: Same value for all transcripts (simple shrinkage)
2. **Gene-level**: Mean of transcript abundances within gene
3. **Long-read informed**: Based on long-read observations

```{r prior-example, eval=FALSE}
# Example: Gene-level prior
compute_gene_level_prior <- function(abundances, transcript_to_gene) {
  gene_means <- tapply(log(abundances + 1e-10), transcript_to_gene, mean)
  prior_means <- gene_means[transcript_to_gene]
  return(prior_means)
}
```


## The EM Algorithm

### Overview

MPAQT uses an Expectation-Maximization (EM) algorithm to maximize the objective function. The algorithm iterates between:

1. **E-step**: Compute expected read assignments
2. **M-step**: Update abundance estimates

### E-Step: Read Assignment

Given current abundances $\boldsymbol{\beta}^{(t)}$, compute the probability that each read in EC $i$ came from transcript $j$:

$$\gamma_{ij}^{(t)} = \frac{\beta_j^{(t)} \cdot P_{ij}}{\sum_{k=1}^{T} \beta_k^{(t)} \cdot P_{ik}}$$

### M-Step: Abundance Update

Update abundances using the Newton-Raphson method. For each transcript $j$:

#### Gradient (First Derivative)

$$\frac{\partial L}{\partial \beta_j} = \underbrace{\sum_{i=1}^{E} \frac{n_i \cdot P_{ij}}{y_i} - \sum_{i=1}^{E} P_{ij}}_{\text{Short-read}} + \underbrace{\frac{c_j}{\beta_j} - p_j}_{\text{Long-read}} - \underbrace{\frac{\lambda (\log\beta_j - \mu_j)}{\beta_j}}_{\text{Prior}}$$

#### Hessian (Second Derivative)

$$\frac{\partial^2 L}{\partial \beta_j^2} = -\sum_{i=1}^{E} \frac{n_i \cdot P_{ij}^2}{y_i^2} - \frac{c_j}{\beta_j^2} - \frac{\lambda(1 - \log\beta_j + \mu_j)}{\beta_j^2}$$

#### Newton-Raphson Update

$$\beta_j^{(t+1)} = \beta_j^{(t)} - \frac{\partial L / \partial \beta_j}{\partial^2 L / \partial \beta_j^2}$$

With bounds to ensure $\beta_j > 0$:

```{r newton-update, eval=FALSE}
# Newton-Raphson update with bounds
newton_update <- function(beta, gradient, hessian, min_beta = 1e-10) {
  # Ensure hessian is negative (concave region)
  if (hessian >= 0) {
    # Use gradient descent fallback
    step <- 0.1 * gradient
  } else {
    step <- -gradient / hessian
  }

  # Apply update with bounds
  new_beta <- beta + step
  new_beta <- max(new_beta, min_beta)

  return(new_beta)
}
```

### Convergence Criteria

The algorithm terminates when:

1. **Relative change**: $\frac{\|\boldsymbol{\beta}^{(t+1)} - \boldsymbol{\beta}^{(t)}\|}{\|\boldsymbol{\beta}^{(t)}\|} < \epsilon$
2. **Likelihood change**: $|L^{(t+1)} - L^{(t)}| < \epsilon$
3. **Maximum iterations**: Reached `max_iter`

```{r convergence, eval=FALSE}
# Check convergence
check_convergence <- function(beta_old, beta_new, tolerance = 1e-4) {
  relative_change <- sqrt(sum((beta_new - beta_old)^2)) / sqrt(sum(beta_old^2))
  return(relative_change < tolerance)
}
```


## Positional Bias Correction

### The Bias Model

RNA degradation and sequencing artifacts cause non-uniform read coverage along transcripts. MPAQT models this with distance-dependent weights.

### 3' Bias Model

For 3' bias (common in poly-A selected data), reads near the 3' end are over-represented:

$$w_j(d) = \exp\left( \sum_{b=1}^{B} \theta_b \cdot \mathbb{1}[d \in \text{bin}_b] \right)$$

where:

- $d$: Distance from 3' end
- $B$: Number of distance bins
- $\theta_b$: Weight for bin $b$

### 5' Bias Model

For 5' bias (common in degraded samples), reads near the 5' end are depleted:

$$w_j(d) = \exp\left( \sum_{b=1}^{B} \phi_b \cdot \mathbb{1}[d \in \text{bin}_b] \right)$$

### Bias Weight Optimization

The bias weights $\boldsymbol{\theta}$ (or $\boldsymbol{\phi}$) are optimized using L-BFGS-B:

$$\min_{\boldsymbol{\theta}} \sum_{i=1}^{E} \left( n_i - \sum_{j=1}^{T} \beta_j \cdot w_j(d_i) \cdot P_{ij} \right)^2$$

```{r bias-optimization, eval=FALSE}
# Optimize bias weights
optimize_bias_weights <- function(counts, P_matrix, abundances,
                                   distances, n_bins = 50) {
  # Create distance bins
  bins <- cut(distances, breaks = n_bins, labels = FALSE)

  # Objective function
  objective <- function(weights) {
    # Apply weights to P matrix
    weighted_P <- P_matrix * exp(weights[bins])

    # Compute expected counts
    expected <- weighted_P %*% abundances

    # Residual sum of squares
    return(sum((counts - expected)^2))
  }

  # Optimize
  result <- optim(
    par = rep(0, n_bins),
    fn = objective,
    method = "L-BFGS-B",
    lower = -5,
    upper = 5
  )

  return(result$par)
}
```


## Prior Integration with Mixed-Effects Models

### The Hierarchical Model

MPAQT can incorporate gene-level information through a mixed-effects model:

$$\log(\beta_{jg}) = \mathbf{X}_{jg}^T \boldsymbol{\theta} + b_g + \epsilon_{jg}$$

where:

- $\beta_{jg}$: Abundance of transcript $j$ in gene $g$
- $\mathbf{X}_{jg}$: Fixed-effect covariates (GC content, length, etc.)
- $\boldsymbol{\theta}$: Fixed-effect coefficients
- $b_g \sim \mathcal{N}(0, \sigma_g^2)$: Gene-level random effect
- $\epsilon_{jg} \sim \mathcal{N}(0, \sigma^2)$: Residual error

### Implementation with GPBoost/lme4

```{r mixed-effects, eval=FALSE}
# Fit mixed-effects model for prior
fit_prior_model <- function(abundances, covariates, gene_ids) {
  # Prepare data
  df <- data.frame(
    log_abundance = log(abundances + 1e-10),
    gc_content = covariates$gc,
    length = log(covariates$length),
    gene = gene_ids
  )

  # Fit with lme4
  model <- lme4::lmer(
    log_abundance ~ gc_content + length + (1 | gene),
    data = df
  )

  # Extract prior parameters
  prior_mean <- predict(model)
  prior_var <- sigma(model)^2

  return(list(mean = prior_mean, variance = prior_var))
}
```


## Two-Phase Quantification

### Pre-Quantification Phase

The pre-quantification phase provides initial abundance estimates and optimizes bias weights:

```{r prequant, eval=FALSE}
# Pre-quantification workflow
prequant_result <- mpaqt_prequant(
  index = idx,
  sr_counts = sr_counts,
  lr_counts = lr_counts,
  positional_bias = "3p",
  n_bins = 50,
  max_iter = 50,        # Fewer iterations
  fit_prior = TRUE,     # Fit prior model
  verbose = TRUE
)
```

**Key outputs**:

- Initial abundance estimates
- Optimized bias weights
- Prior model parameters
- Convergence diagnostics

### Post-Quantification Phase

The post-quantification phase refines estimates using the pre-computed parameters:

```{r postquant, eval=FALSE}
# Post-quantification workflow
result <- mpaqt_postquant(
  index = idx,
  sr_counts = sr_counts,
  lr_counts = lr_counts,
  prequant = prequant_result,
  max_iter = 100,
  tolerance = 1e-6,     # Tighter convergence
  compute_uncertainty = TRUE,
  verbose = TRUE
)
```

**Key outputs**:

- Final abundance estimates (TPM)
- Uncertainty estimates (if requested)
- Convergence information
- Log-likelihood values


## Uncertainty Estimation

### Fisher Information

Uncertainty in abundance estimates is computed using the observed Fisher information:

$$I(\beta_j) = -\frac{\partial^2 L}{\partial \beta_j^2}$$

The standard error is:

$$SE(\beta_j) = \frac{1}{\sqrt{I(\beta_j)}}$$

### Confidence Intervals

Approximate 95% confidence intervals:

$$\left[ \beta_j \cdot e^{-1.96 \cdot SE(\log\beta_j)}, \quad \beta_j \cdot e^{1.96 \cdot SE(\log\beta_j)} \right]$$

```{r uncertainty, eval=FALSE}
# Compute uncertainty from Fisher information
compute_uncertainty <- function(abundances, hessian_diag) {
  # Fisher information (negative Hessian)
  fisher_info <- -hessian_diag

  # Standard errors
  se <- 1 / sqrt(pmax(fisher_info, 1e-10))

  # Confidence intervals (log scale)
  ci_lower <- abundances * exp(-1.96 * se / abundances)
  ci_upper <- abundances * exp(1.96 * se / abundances)

  return(list(se = se, ci_lower = ci_lower, ci_upper = ci_upper))
}
```


## Workflow Diagram

The complete MPAQT workflow:

```
                    +-------------------+
                    |   Reference Files |
                    | (FASTA + GTF)     |
                    +--------+----------+
                             |
                             v
                    +-------------------+
                    |   mpaqt_index()   |
                    | - Build kallisto  |
                    | - Compute P matrix|
                    +--------+----------+
                             |
            +----------------+----------------+
            |                                 |
            v                                 v
+---------------------+            +---------------------+
| Short-Read FASTQ    |            | Long-Read FASTQ     |
+----------+----------+            +----------+----------+
           |                                  |
           v                                  v
+---------------------+            +---------------------+
| mpaqt_prepare_      |            | mpaqt_prepare_      |
| short_reads()       |            | long_reads()        |
| - kallisto bus      |            | - minimap2          |
| - Extract EC counts |            | - Bambu             |
+----------+----------+            +----------+----------+
           |                                  |
           v                                  v
+---------------------+            +---------------------+
| mpaqt_counts_sr     |            | mpaqt_counts_lr     |
| (EC count vector)   |            | (Transcript counts) |
+----------+----------+            +----------+----------+
           |                                  |
           +----------------+----------------+
                            |
                            v
               +-------------------------+
               |   mpaqt_prequant()      |
               | - Initial EM            |
               | - Bias optimization     |
               | - Prior model fitting   |
               +------------+------------+
                            |
                            v
               +-------------------------+
               |   mpaqt_postquant()     |
               | - Final EM              |
               | - Uncertainty estimation|
               +------------+------------+
                            |
                            v
               +-------------------------+
               |   mpaqt_quant_result    |
               | - abundances (TPM)      |
               | - uncertainty (SE)      |
               | - convergence info      |
               +-------------------------+
```


## Algorithm Summary

### Key Equations Reference

| Component | Equation |
|-----------|----------|
| **Objective** | $L(\boldsymbol{\beta}) = L_{SR} + L_{LR} - R$ |
| **SR Likelihood** | $L_{SR} = \sum_i [n_i \log(y_i) - y_i]$ |
| **Expected EC Count** | $y_i = \sum_j \beta_j P_{ij}$ |
| **LR Likelihood** | $L_{LR} = \sum_j [c_j \log(\beta_j p_j) - \beta_j p_j]$ |
| **Prior** | $R = \frac{\lambda}{2} \sum_j (\log\beta_j - \mu_j)^2$ |
| **E-step** | $\gamma_{ij} = \frac{\beta_j P_{ij}}{\sum_k \beta_k P_{ik}}$ |
| **Newton Update** | $\beta_j^{new} = \beta_j - \frac{\nabla L}{\nabla^2 L}$ |

### Default Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `max_iter` | 100 | Maximum EM iterations |
| `tolerance` | 1e-4 | Convergence threshold |
| `lambda` | 0.1 | Prior strength |
| `n_bins` | 50 | Bias correction bins |
| `prior_start` | 25 | Iteration to enable prior |
| `convergence_start` | 25 | Iteration to check convergence |


## References

1. **kallisto**: Bray, N.L., et al. (2016). Near-optimal probabilistic RNA-seq quantification. Nature Biotechnology.

2. **EM Algorithm for RNA-seq**: Pachter, L. (2011). Models for transcript quantification from RNA-Seq. arXiv.

3. **bambu**: Chen, Y., et al. (2021). Bambu: context-aware transcript quantification from long read RNA-seq experiments.

4. **Mixed-Effects Models**: Bates, D., et al. (2015). Fitting Linear Mixed-Effects Models Using lme4. Journal of Statistical Software.

5. **Fisher Information**: Efron, B. & Hinkley, D.V. (1978). Assessing the accuracy of the maximum likelihood estimator.


## Session Information

```{r session-info, eval=FALSE}
sessionInfo()
```
