---
title: "MPAQT Quantification: Implemented Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MPAQT Quantification: Implemented Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

## Scope

MPAQT integrates short-read equivalence-class counts with optional long-read
transcript counts, positional-bias weights, UMI normalization, and
transcript-level priors. This article describes the algorithm implemented in
package version 2.4.0.

The internal function is named `run_em_algorithm()`, but the implementation is
not a classical EM routine that materializes an E-step assignment matrix.
Within each iteration it performs sequential transcript-wise coordinate Newton
updates and immediately updates the expected equivalence-class counts.

## Notation and Short-Read Model

Let:

- $n_i$ be the observed count for short-read equivalence class (EC) $i$;
- $\beta_j \ge 0$ be the raw abundance for transcript $j$;
- $P_{ij}$ be the stored EC-by-transcript design/exposure value; and
- $y_i = \sum_j P_{ij}\beta_j$ be the expected EC count.

For a default bulk index, the nonzero `x` values in `index$p_matrices[[j]]`
are counts produced by simulated reads, not columns that are automatically
normalized to sum to one. Each per-transcript table stores the EC indices `i`,
the values `x`, and the associated 5' and 3' distances.

Ignoring constants, the short-read Poisson negative log-likelihood is

$$
\operatorname{NLL}_{SR}(\boldsymbol\beta)
= \sum_i \left[y_i - n_i\log(y_i)\right].
$$

### Optional UMI normalization

`do_umi_correction = TRUE` normalizes each transcript's P column to sum to one.
Without positional weights, or with `umi_correction_timing = "pre"`, this is
done before the coordinate iterations. With positional weights and
`umi_correction_timing = "post"`, MPAQT multiplies by the positional weights
first and normalizes the resulting weighted column during each iteration.

This P-matrix normalization is distinct from UMI molecule deduplication in
`mpaqt_prepare_short_reads_sc()`. Read preparation defaults to
`do_umi_dedup = FALSE`; users who want molecule-level EC counts must enable
deduplication explicitly.

## Sequential Coordinate Updates

MPAQT initializes all $\beta_j$ and $y_i$ to zero. It then visits transcripts
in index order during every iteration.

### First iteration

Because the Poisson expressions are unstable at an all-zero initialization,
the first iteration uses a Gaussian residual approximation. For the current
transcript column $p_j$,

$$
\Delta_j =
\frac{\sum_i (n_i-y_i)P_{ij}}
     {\sum_i P_{ij}^2}.
$$

### Later iterations

For subsequent iterations, with a small numerical constant $\epsilon$, the
implemented short-read gradient and positive curvature terms are

$$
g_j = \sum_i P_{ij}
\left(1-\frac{n_i+\epsilon}{y_i+\epsilon}\right)
$$

and

$$
h_j = \sum_i (n_i+\epsilon)
\left(\frac{P_{ij}}{y_i+\epsilon}\right)^2.
$$

The unregularized coordinate step is $\Delta_j=-g_j/h_j$. Non-finite steps
are replaced with zero, and the lower bound
$\Delta_j \ge -\beta_j+10^{-10}$ preserves positive abundances. After each
step, the implementation immediately applies

$$
\beta_j \leftarrow \beta_j+\Delta_j, \qquad
y_i \leftarrow y_i+P_{ij}\Delta_j.
$$

Consequently, later transcripts in an iteration see the updates already made
to earlier transcripts. MPAQT does not construct or store latent assignment
probabilities $\gamma_{ij}$.

## Log-Scale Regularization

Starting after the first iteration, positive transcript abundances receive a
quadratic penalty in log space:

$$
R_j = \frac{(\log\beta_j-\mu_j)^2}{2\sigma^2}.
$$

The penalty contributions added to the coordinate gradient and curvature are

$$
g_{R,j}=\frac{\log\beta_j-\mu_j}{\beta_j\sigma^2}
$$

and

$$
h_{R,j}=\frac{-\log\beta_j+\mu_j+1}{\beta_j^2\sigma^2}.
$$

The regularized Newton step is bounded relative to the unregularized step and
the step toward the prior mode, $\exp(\mu_j)-\beta_j$. When it lies outside
that interval, MPAQT minimizes the one-dimensional penalized Poisson objective
with `stats::optimize()` over the bounded interval.

Without a user prior, $\mu$ and $\sigma^2$ are updated from the mean and mean
squared deviation of positive log abundances after each iteration. The
variance is bounded below by $10^{-6}$ in post-quantification.

## Long-Read Integration

When long-read counts are present, MPAQT fits a Poisson GLM after every
coordinate sweep:

$$
c_j \sim \operatorname{Poisson}(\beta_j q_j), \qquad
q_j=\exp(C_j\boldsymbol\theta).
$$

Here $c_j$ is the long-read transcript count and $C_j$ is the row of index
covariates. The current index covariates are an intercept, log GC ratio, log
transcript length, and a protein-coding indicator. Sequencing-depth scaling is
absorbed by the fitted intercept. Although the result field is named
`lr_coverage_probs`, $q_j$ is a positive rate/exposure multiplier and is not
constrained to the interval $[0,1]$.

During an abundance coordinate update, a long-read observation is appended
only when $c_j>0$. A transcript with zero long-read counts therefore does not
receive a direct zero-count long-read coordinate term. Zero counts still
participate in the GLM fit and in the diagnostic likelihood described below.

## Prior Specifications

`prior_model` accepts the following forms:

| Value | Implemented prior mean behavior |
|-------|---------------------------------|
| `NULL` | Uses the global mean of positive log abundances |
| `"shrinkage"` | Uses transcript prior predictions initialized to zero from `prior_start` |
| `"long_read"` | Uses `log(long_read_count + 1)` as transcript prior predictions |
| data frame/data table | Requires `transcript_id`; remaining columns are GPBoost covariates |
| list with `mean` | Uses a full-length or transcript-named custom mean vector |

For a data-frame prior, transcript versions are trimmed when matching the
covariate table to the index. Beginning at `prior_start`, GPBoost is fitted
after calculation of the diagnostic likelihood and after the convergence
check. Its predictions and variance therefore affect the next iteration.

To ensure a data-frame prior is used, choose `convergence_start` later than
`prior_start`. The single-cell GTEx example uses:

```{r prior-schedule, eval=FALSE}
results <- mpaqt_quant_sc(
  index = index,
  sr_counts_list = sr_list,
  prior_model = prior_data,
  prior_start = 25L,
  convergence_start = 50L
)
```

## Positional-Bias Prequantification

Positional-bias correction is optional. When `positional_bias` is `"3p"` or
`"5p"`, `mpaqt_quant()` first calls `mpaqt_prequant()`.

The selected transcript-EC distances $d_{ij}$ are divided into global quantile
bins. All bin weights start at one. After the default 20-iteration warm-up,
MPAQT builds an EC-by-bin expected-count matrix $S$ and minimizes

$$
\operatorname{NLL}_{bias}(\mathbf w)
=\sum_i\left[\lambda_i-n_i\log(\lambda_i)\right],
\qquad
\boldsymbol\lambda=S\exp(\log\mathbf w),
$$

using L-BFGS-B. At each update, log weights are bounded to the current log
weights plus or minus one. For identifiability, fitted weights are divided by
the first-bin weight and abundances are multiplied by that same value.

The `"3p"` and `"5p"` settings select the distance origin; they do not impose a
monotonic enrichment direction. `mpaqt_prequant()` returns positional weights,
quantile boundaries, bias metadata, convergence information, and initial
abundances. It does not accept long-read counts or fit a transcript prior.

## Diagnostic Likelihood and Convergence

The scalar stored as `log_likelihood` is a convergence diagnostic. It is not
the complete penalized objective used by the coordinate steps.

The diagnostic begins with

$$
D_{SR}=\sum_i[n_i\log(y_i+\epsilon)-y_i]
$$

and adds $-T\log(\sigma^2)/2$. It does not include the transcript-specific
quadratic terms $(\log\beta_j-\mu_j)^2/(2\sigma^2)$.

When long reads are present, the implementation also adds

$$
\sum_j[c_j\log(\beta_jq_j+\epsilon)-\beta_jq_j]
$$

and the fitted GLM's `logLik()` value. These contributions should therefore be
interpreted as the current diagnostic calculation, not as a single separately
defined likelihood optimized exactly by every coordinate update.

After `convergence_start`, the algorithm declares convergence only when the
diagnostic change is positive and smaller than `tolerance`:

$$
0 < D^{(t)}-D^{(t-1)} < \texttt{tolerance}.
$$

There is no separate relative-abundance convergence criterion.

## Two-Phase API

The main `mpaqt_quant()` wrapper runs prequantification only when positional
bias is requested. The same work can be run explicitly:

```{r two-phase, eval=FALSE}
prequant <- mpaqt_prequant(
  index = index,
  sr_counts = sr_counts,
  positional_bias = "3p",
  n_bins = 50L
)

mpaqt_save_prequant(prequant, "results/mpaqt.prequant.rds")

result <- mpaqt_postquant(
  index = index,
  sr_counts = sr_counts,
  lr_counts = lr_counts,
  prequant = prequant,
  prior_model = prior_data,
  prior_start = 25L,
  convergence_start = 50L,
  compute_uncertainty = TRUE
)
```

Long-read observations and priors belong to `mpaqt_postquant()`; they are not
inputs or outputs of `mpaqt_prequant()`.

## Uncertainty

With `compute_uncertainty = TRUE`, MPAQT returns a diagonal, short-read-only
approximation. For each positive transcript abundance,

$$
I_j=\sum_i\frac{P_{ij}^2n_i}{\max(y_i,10^{-10})^2},
\qquad
SE_j=I_j^{-1/2}.
$$

This calculation does not include off-diagonal transcript covariance,
long-read information, prior curvature, or explicit positional-weight terms.
It returns standard errors only; the package does not currently return
confidence intervals.

## Returned Abundances and TPM

An `mpaqt_quant_result` stores raw coordinate estimates in `abundances` and a
separately scaled vector in `tpm`:

$$
\operatorname{TPM}_j=10^6\frac{\beta_j}{\sum_k\beta_k}.
$$

Use `abundances(result)` and `tpm(result)` to keep these values distinct. The
`tpm` field is calculated for every `normalize` setting. `normalize = "depth"`
also stores the million-scaled values in `normalized_counts`; `"tpm"` and
`"none"` retain raw abundances in that field.

## Defaults by Entry Point

| Parameter | `mpaqt_quant()` | `mpaqt_prequant()` | `mpaqt_postquant()` | `mpaqt_quant_sc()` |
|-----------|-----------------|--------------------|---------------------|--------------------|
| `max_iter` | 100 | 100 | 100 | 100 |
| `tolerance` | `1e-4` | `1e-1` | `1e-4` | `1e-4` |
| `n_bins` | 50 | 50 | from prequant | 50 |
| `weight_update_start` | internal prequant default | 20 | not applicable | internal prequant default |
| `prior_start` | 25 | not applicable | 50 | 50 |
| `convergence_start` | 25 | 25 | 25 | 25 |
| `do_umi_correction` | `FALSE` | not applicable | `FALSE` | `TRUE` |
| `umi_correction_timing` | `"post"` | not applicable | `"post"` | `"post"` |
| `compute_uncertainty` | `FALSE` | not applicable | `FALSE` | not exposed (`FALSE`) |

`mpaqt_quant()` passes its public post-quantification defaults to
`mpaqt_postquant()`, which is why its `prior_start` default differs from a
direct call to `mpaqt_postquant()`.

## References

1. Bray, N. L., et al. (2016). Near-optimal probabilistic RNA-seq
   quantification. *Nature Biotechnology*.
2. Chen, Y., et al. (2021). Context-aware transcript quantification from long
   read RNA-seq data with bambu. *Nature Methods*.
3. Sigrist, F. (2022). Gaussian process boosting. *Journal of Machine Learning
   Research*.
