---
title: "Single-Cell RNA-seq Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-Cell RNA-seq Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

This tutorial builds one cluster-level single-cell RNA-seq analysis through
four cumulative configurations:

1. short reads;
2. short and long reads;
3. short and long reads with positional-bias correction; and
4. short and long reads with positional-bias correction and a GTEx prior.

Each configuration reuses the objects prepared in the previous sections, so
the effect of each added source of information can be compared directly.

## What This Tutorial Produces

By the end, you will have:

- one `mpaqt_index` object for the reference transcriptome;
- cluster-level short-read and long-read count lists;
- four named lists of `mpaqt_quant_result` objects; and
- a transcript-by-cluster TPM table for downstream analysis.

The four result lists are named `sr_results`, `sr_lr_results`,
`sr_lr_bias_results`, and `sr_lr_bias_prior_results`.

## Required Inputs

The short-read configuration requires:

- a GTF annotation and matching transcriptome FASTA;
- paired 10x Chromium FASTQ files, with barcode and UMI information in R1 and
  cDNA sequence in R2;
- a cluster-assignment CSV; and
- `kallisto` and `bustools` available on your system.

The cluster file must have barcodes in its first column and cluster labels in
its second column. Additional columns are ignored. For example:

```text
barcode,cluster
AAACCCAAGAAACACT-1,Astrocyte
AAACCCAAGAAACCAT-1,Neuron
AAACCCAAGAAACCCG-1,Neuron
```

The long-read configuration additionally requires a transcript-by-cell count
matrix. Its first column contains transcript IDs, and its remaining column
names are cell barcodes:

```text
transcript_id,AAACCCAAGAAACACT-1,AAACCCAAGAAACCAT-1
ENST00000456328.2,2,0
ENST00000450305.2,0,3
```

The final configuration also uses the GTEx PCA table downloaded in
[Add the GTEx Prior](#add-the-gtex-prior).

## Create or Load the Index

Load MPAQT and create the index once for a reference:

```{r sc-create-index}
library(mpaqt)

index <- mpaqt_index(
    annotation = "reference/gencode.annotation.gtf",
    transcriptome = "reference/gencode.transcripts.fa",
    output_file = "reference/mpaqt.index.rds",
    read_length = 91L,
    threads = 8L
)
```

Set `read_length` to the length of the cDNA sequence represented by R2. Reuse
the saved index in later runs:

```{r sc-load-index}
index <- mpaqt_read_index("reference/mpaqt.index.rds")
```

See the [index reference](../reference/mpaqt_index.html) for other supported
reference inputs.

## Prepare Single-Cell Short Reads

Prepare cluster-level equivalence-class counts from the 10x FASTQ files and
cluster assignments:

```{r sc-prepare-short}
sr_list <- mpaqt_prepare_short_reads_sc(
    index = index,
    fastq_1 = "data/sample_R1.fastq.gz",
    fastq_2 = "data/sample_R2.fastq.gz",
    clusters_file = "data/clusters.csv",
    do_umi_dedup = TRUE,
    umi_dedup_mode = "fractional_proportional",
    output_dir = "results/short-read",
    technology = "10xv3",
    threads = 8L
)
```

`sr_list` contains one `mpaqt_counts_sr` object per cluster and an
`"all_clusters"` object that pools their counts. With the settings above, each
barcode-UMI molecule contributes a total count of one while its evidence can
be distributed across equivalence classes.

UMI handling occurs at two separate stages. `do_umi_dedup = TRUE` collapses
read records to molecule-level counts during preparation. Later,
`mpaqt_quant_sc()` uses `do_umi_correction = TRUE` by default to normalize the
short-read probability matrices during quantification.

## Quantify with Short Reads

Run the first configuration for every entry in `sr_list`:

```{r sc-quant-short}
sr_results <- mpaqt_quant_sc(
    index = index,
    sr_counts_list = sr_list
)
```

The returned list uses the same cluster names as `sr_list` and also includes
the pooled `"all_clusters"` result.

## Prepare Single-Cell Long Reads

Aggregate the transcript-by-cell long-read matrix using the same cluster file:

```{r sc-prepare-long}
lr_list <- mpaqt_prepare_long_reads_sc(
    index = index,
    count_file = "data/long_read_counts.csv",
    clusters_file = "data/clusters.csv",
    output_dir = "results/long-read"
)
```

`lr_list` contains one `mpaqt_counts_lr` object per cluster plus an
`"all_clusters"` object. The short- and long-read lists must use the same
cluster names. Cell barcodes in the count-matrix columns must match those in
the cluster file exactly.

## Quantify with Short and Long Reads

Add the long-read counts to obtain the second configuration:

```{r sc-quant-short-long}
sr_lr_results <- mpaqt_quant_sc(
    index = index,
    sr_counts_list = sr_list,
    lr_counts_list = lr_list
)
```

This call reuses the same cluster-level short-read evidence and adds
transcript-level long-read evidence for each cluster.

## Add Positional-Bias Correction

For a 3'-biased protocol or observed 3' enrichment, add 3' positional-bias
correction to obtain the third configuration:

```{r sc-quant-bias}
sr_lr_bias_results <- mpaqt_quant_sc(
    index = index,
    sr_counts_list = sr_list,
    lr_counts_list = lr_list,
    positional_bias = "3p"
)
```

MPAQT estimates one set of positional weights from the pooled
`sr_list[["all_clusters"]]` counts and applies those weights to each cluster.
For a 5'-biased protocol or observed 5' enrichment, use
`positional_bias = "5p"` instead.

## Add the GTEx Prior

The GTEx prior table contains transcript expression embeddings derived from
principal component analysis of GTEx v8 multi-tissue bulk RNA-seq data. Each
row represents one transcript; the `transcript_id` column is followed by 200
features named `PC1` through `PC200`. MPAQT uses these scores as predictive
features when fitting its abundance prior.

Download [`gtex_PCA_output.csv.gz`](https://zenodo.org/records/21478011/files/gtex_PCA_output.csv.gz?download=1)
from the [Zenodo record](https://zenodo.org/records/21478011) and decompress it:

```{bash sc-download-prior}
wget -O gtex_PCA_output.csv.gz "https://zenodo.org/records/21478011/files/gtex_PCA_output.csv.gz?download=1"
gunzip gtex_PCA_output.csv.gz
```

Read the decompressed table, then pass it directly to `prior_model` in the
fourth configuration:

```{r sc-quant-prior}
gtex_prior_path <- "gtex_PCA_output.csv"
prior_data <- data.table::fread(gtex_prior_path)

sr_lr_bias_prior_results <- mpaqt_quant_sc(
    index = index,
    sr_counts_list = sr_list,
    lr_counts_list = lr_list,
    positional_bias = "3p",
    prior_model = prior_data,
    prior_start = 25L,
    convergence_start = 50L
)
```

MPAQT matches `prior_data$transcript_id` to the index after removing version
suffixes. With these iteration settings, prior fitting begins at iteration 25
and convergence is not checked until iteration 50, allowing the fitted prior
to affect the abundance estimates before the run can stop.

These four configurations are cumulative to make their comparison clear. The
R API does not make long reads, positional-bias correction, and the prior
intrinsically dependent: arguments can be omitted when a study calls for a
different combination.

## Inspect, Combine, and Save Results

Choose the result list for the configuration you want to take forward. Here we
use the fourth configuration and exclude the pooled `"all_clusters"` result
from the cluster-by-cluster output:

```{r sc-inspect-results}
final_results <- sr_lr_bias_prior_results
cluster_ids <- setdiff(names(final_results), "all_clusters")
cluster_results <- final_results[cluster_ids]

head(tpm(cluster_results[[1]]))

combined_tpm <- do.call(
    combine_quant_results,
    c(
        unname(cluster_results),
        list(sample_names = cluster_ids)
    )
)

head(combined_tpm)
```

`combined_tpm` has one `transcript_id` column followed by one TPM column per
cluster. Save the per-cluster result objects, per-cluster TPM tables, and the
combined table together:

```{r sc-save-results}
mpaqt_save_result_sc(
    cluster_results,
    output_dir = "results/quantification",
    prefix = "sample"
)
```

Use `sr_results`, `sr_lr_results`, or `sr_lr_bias_results` in the same way when
you want to inspect or save an earlier configuration.

## Alternative Input Formats

If kallisto has already produced matching BUS and equivalence-class files,
prepare the single-cell short-read list without rerunning pseudoalignment:

```{r sc-short-bus}
sr_list <- mpaqt_prepare_short_reads_sc(
    index = index,
    clusters_file = "data/clusters.csv",
    bus_file = "data/output.bus",
    ec_file = "data/matrix.ec",
    do_umi_dedup = TRUE,
    umi_dedup_mode = "fractional_proportional",
    output_dir = "results/short-read",
    technology = "10xv3",
    threads = 8L
)
```

Previously prepared cluster objects can be reconstructed into named lists from
their stored `sample_id` values:

```{r sc-load-rds}
sr_files <- list.files(
    "results/short-read",
    pattern = "\\.short_read\\.rds$",
    full.names = TRUE
)
sr_list <- lapply(sr_files, mpaqt_read_short_read_counts)
names(sr_list) <- vapply(sr_list, function(x) x$sample_id, character(1))

lr_files <- list.files(
    "results/long-read",
    pattern = "\\.long_read\\.rds$",
    full.names = TRUE
)
lr_list <- lapply(lr_files, mpaqt_read_long_read_counts)
names(lr_list) <- vapply(lr_list, function(x) x$sample_id, character(1))
```

See the [single-cell short-read](../reference/mpaqt_prepare_short_reads_sc.html)
and [single-cell long-read](../reference/mpaqt_prepare_long_reads_sc.html)
references for complete input details.

## Troubleshooting

**Cluster names differ between lists.** Prepare short and long reads from the
same cluster file, or rename list entries so every short-read cluster has the
corresponding long-read entry.

**No long-read barcodes match a cluster.** Compare the matrix column names with
the first column of `clusters.csv`, including any `-1` suffixes.

**Transcript IDs do not match.** Use the same reference release for the index
and long-read counts. The GTEx table can be matched with or without transcript
version suffixes.

**kallisto, bustools, or GPBoost is unavailable.** Follow the dependency checks
in the [installation article](installation.html), then restart R if your
environment changed.

**The final result contains `all_clusters`.** This pooled entry is useful for
shared weight estimation. Remove it with `setdiff()` as shown before combining
or saving per-cluster results.

## Next Steps

- See the [bulk workflow](bulk-workflow.html) for its two tutorial
  configurations.
- Read [quantification methods](quantification-methods.html) for the statistical
  model and algorithm details.
- Use the [function reference](../reference/index.html) for complete argument
  descriptions.
