---
title: "Single-Cell RNA-seq Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-Cell RNA-seq Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article documents the MPAQT workflow for single-cell RNA-seq transcript quantification. The single-cell workflow is designed for cluster-level quantification, aggregating cells by cluster assignments.

## Overview

Single-cell RNA-seq presents unique challenges for transcript-level quantification:

- **Sparse data**: Individual cells have limited coverage
- **Technical noise**: High dropout rates and amplification bias
- **Cell heterogeneity**: Different cell types express different isoforms

MPAQT addresses these challenges by:

1. **Cluster aggregation**: Pooling reads from cells in the same cluster
2. **Shared bias correction**: Learning positional bias from combined data
3. **Per-cluster quantification**: Running EM algorithm per cluster

```
SINGLE-CELL WORKFLOW
====================

┌─────────────────────┐     ┌─────────────────────┐
│     FASTQ Files     │     │  Cluster Assignments│
│  (10x Chromium)     │     │   (barcode→cluster) │
└──────────┬──────────┘     └──────────┬──────────┘
           │                           │
           └─────────────┬─────────────┘
                         ▼
           ┌─────────────────────────┐
           │ mpaqt_prepare_short_    │
           │      reads_sc()         │
           │                         │
           │ Groups reads by cluster │
           └─────────────┬───────────┘
                         │
                         ▼
           ┌─────────────────────────┐
           │    mpaqt_quant_sc()     │
           │                         │
           │ Quantifies each cluster │
           │ independently           │
           └─────────────┬───────────┘
                         │
                         ▼
           ┌─────────────────────────┐
           │  Per-cluster TPM +      │
           │  Combined matrix        │
           └─────────────────────────┘
```

---

## Progressive Workflow Examples

This section demonstrates MPAQT single-cell capabilities through progressively complex examples. Each level shows **CLI first (where applicable), then R API**.

### Level 1: Short-Read Only

The simplest single-cell MPAQT workflow processes 10x Chromium data by cluster.

#### Using CLI

```bash
# Process 10x Chromium v3 data
mpaqt prepare-sr my_index.rds \
  --fastq-r1 sample_R1.fastq.gz \
  --fastq-r2 sample_R2.fastq.gz \
  --technology 10xv3 \
  --output-dir results/ \
  --threads 8

# Quantify by cluster
mpaqt quant-sc my_index.rds results/mpaqt.sr.rds \
  --clusters clusters.csv \
  --output-dir results/
```

#### Using R API

```{r sc-level1, eval=FALSE}
library(mpaqt)

# Load or create index (same as bulk)
index <- mpaqt_read_index("my_index.rds")

# Process single-cell short reads by cluster
sr_list <- mpaqt_prepare_short_reads_sc(
  index = index,
  fastq_1 = "sample_R1.fastq.gz",
  fastq_2 = "sample_R2.fastq.gz",
  clusters_file = "clusters.csv",
  technology = "10xv3",
  output_dir = "results/",
  threads = 8
)

# Quantify all clusters
results <- mpaqt_quant_sc(
  index = index,
  sr_counts_list = sr_list
)

# Get combined TPM matrix (transcripts × clusters)
tpm_matrix <- combine_quant_results(results)
```

> **Note:** `mpaqt_quant_sc()` now enables `do_umi_correction = TRUE` by default. This renormalizes each transcript P matrix before post-quantification, which is appropriate for UMI-based single-cell data. Set `do_umi_correction = FALSE` only if you explicitly want the bulk-style behavior.

---

### Level 2: Adding Positional Bias Correction

In single-cell quantification, positional bias correction uses the combined "all_clusters" data for robust weight estimation, then applies these weights to each cluster.

#### Using R API

```{r sc-level2, eval=FALSE}
results <- mpaqt_quant_sc(
  index = index,
  sr_counts_list = sr_list,
  positional_bias = "3p",
  n_bins = 50
)
```

> **Note:** Bias weights are learned from pooled data across all clusters, then applied per-cluster. This provides more robust estimation than learning bias separately per cluster.

---

### Level 3: With Long-Read Integration (Advanced)

If you have matching single-cell long-read data (e.g., MAS-ISO-seq), you can integrate it for improved accuracy.

#### Using R API

```{r sc-level3, eval=FALSE}
# Prepare SC long-read counts (cluster-level)
lr_list <- mpaqt_prepare_long_reads_sc(
  index = index,
  counts_file = "lr_counts_by_cluster.csv",
  clusters_file = "clusters.csv",
  output_dir = "results/"
)

# Quantify with both data types
results <- mpaqt_quant_sc(
  index = index,
  sr_counts_list = sr_list,
  lr_counts_list = lr_list,
  positional_bias = "3p"
)

# Save results
mpaqt_save_result_sc(results, "results/")
```

> **Note:** Single-cell long-read integration is less common than bulk. If you don't have cluster-level long-read data, use Level 1 or 2.

---

## Detailed Reference

The following sections provide comprehensive documentation for each stage.

---

## Stage 1: Index Creation

The index creation process is identical to bulk RNA-seq. See the [Bulk Workflow](bulk-workflow.html) for details.

```r
library(mpaqt)

# Create index (same as bulk)
index <- mpaqt_index(
  annotation = "reference/gencode.v44.annotation.gtf",
  transcriptome = "reference/gencode.v44.transcripts.fa",
  output_file = "reference/mpaqt.index.rds",
  read_length = 91,  # Typical 10x read length
  threads = 8
)
```

---

## Stage 2: Single-Cell Short-Read Processing

### Input Requirements

1. **FASTQ files**: 10x Chromium output (R1 contains barcode+UMI, R2 contains cDNA)
2. **Cluster file**: CSV mapping barcodes to cluster assignments

#### Cluster File Format

```
barcode,cluster
AAACCTGAGAAACCAT-1,1
AAACCTGAGAAACCGC-1,1
AAACCTGAGAAACCTA-1,2
AAACCTGAGAAACGAG-1,3
...
```

### Input Pathways

```
SINGLE-CELL SHORT-READ PATHWAYS
===============================

Pathway A: Raw FASTQ + Clusters (Full pipeline)
───────────────────────────────────────────────
  mpaqt_prepare_short_reads_sc(
      index         = index,
      fastq_1       = "sample_R1.fastq.gz",  # Barcode+UMI
      fastq_2       = "sample_R2.fastq.gz",  # cDNA
      clusters_file = "clusters.csv",
      technology    = "10xv3"
  )

Pathway B: Pre-computed BUS + EC + Clusters
───────────────────────────────────────────
  mpaqt_prepare_short_reads_sc(
      index         = index,
      bus_file      = "kallisto_out/output.bus",
      ec_file       = "kallisto_out/matrix.ec",
      clusters_file = "clusters.csv"
  )

Pathway C: Pre-computed Cluster RDS Files
─────────────────────────────────────────
  mpaqt_prepare_short_reads_sc(
      index   = index,
      rds_dir = "previous_run/clusters/"
  )
```

### Function Reference: `mpaqt_prepare_short_reads_sc()`

```r
mpaqt_prepare_short_reads_sc(
  index,                  # mpaqt_index object
  fastq_1 = NULL,         # Read 1 FASTQ (barcode + UMI)
  fastq_2 = NULL,         # Read 2 FASTQ (cDNA)
  clusters_file = NULL,   # Barcode to cluster mapping
  bus_file = NULL,        # Pre-computed BUS file
  ec_file = NULL,         # Pre-computed EC file
  rds_dir = NULL,         # Directory with cluster RDS files
  output_dir = tempdir(),
  technology = "10xv3",   # 10xv2, 10xv3, or 10xv4
  threads = 1L,
  verbose = TRUE
)
```

### Supported Technologies

| Technology | Barcode | UMI | Description |
|------------|---------|-----|-------------|
| `10xv2` | 16 bp | 10 bp | 10x Chromium v2 |
| `10xv3` | 16 bp | 12 bp | 10x Chromium v3 |
| `10xv4` | 16 bp | 12 bp | 10x Chromium v4 |

### Example: Process Single-Cell Short Reads

```r
# From 10x Chromium FASTQ files
sr_list <- mpaqt_prepare_short_reads_sc(
  index = index,
  fastq_1 = "data/sample_R1.fastq.gz",
  fastq_2 = "data/sample_R2.fastq.gz",
  clusters_file = "data/clusters.csv",
  technology = "10xv3",
  output_dir = "results/",
  threads = 8
)

# Returns named list of mpaqt_counts_sr objects
names(sr_list)
# [1] "1" "2" "3" "4" "5" "all_clusters"

# "all_clusters" contains combined counts for bias estimation
print(sr_list[["1"]])
# mpaqt_counts_sr object (cluster: 1)
#   Equivalence classes: 234,567
#   Total counts: 3,456,789
#   Cells in cluster: 1,234
```

### Output Structure

The function returns a named list where:

- Each cluster ID maps to an `mpaqt_counts_sr` object
- A special `"all_clusters"` entry contains combined counts

Files saved to `output_dir`:
```
results/
├── cluster_1.short_read.rds
├── cluster_2.short_read.rds
├── cluster_3.short_read.rds
├── ...
└── all_clusters.short_read.rds
```

---

## Stage 3: Single-Cell Long-Read Processing (Optional)

Long-read single-cell sequencing (e.g., MAS-ISO-seq) can provide isoform information. Due to lower throughput, long-read data is typically aggregated at the cluster level.

### Function Reference: `mpaqt_prepare_long_reads_sc()`

```r
mpaqt_prepare_long_reads_sc(
  index,                  # mpaqt_index object
  count_file = NULL,      # Transcript x barcode count matrix
  clusters_file = NULL,   # Barcode to cluster mapping
  rds_dir = NULL,         # Directory with cluster RDS files
  output_dir = tempdir(),
  verbose = TRUE
)
```

### Example: Process Single-Cell Long Reads

```r
# From count matrix + cluster assignments
lr_list <- mpaqt_prepare_long_reads_sc(
  index = index,
  count_file = "data/lr_counts_matrix.csv",
  clusters_file = "data/clusters.csv",
  output_dir = "results/"
)

# Returns named list of mpaqt_counts_lr objects
names(lr_list)
# [1] "1" "2" "3" "4" "5"
```

### Count File Format

The long-read count file should have transcript IDs as the first column and barcodes as column headers:

```
transcript_id,AAACCTGAGAAACCAT-1,AAACCTGAGAAACCGC-1,AAACCTGAGAAACCTA-1
ENST00000456328.2,5,0,2
ENST00000450305.2,12,8,0
ENST00000488147.1,0,3,1
...
```

---

## Stage 4: Single-Cell Quantification

The `mpaqt_quant_sc()` function processes all clusters efficiently:

1. **Shared pre-quantification**: Bias weights estimated from combined data
2. **Per-cluster EM**: Each cluster quantified independently
3. **Combined output**: TPM matrix across all clusters

### Function Reference: `mpaqt_quant_sc()`

```r
mpaqt_quant_sc(
  index,                      # mpaqt_index object
  sr_counts_list,             # Named list of mpaqt_counts_sr
  lr_counts_list = NULL,      # Named list of mpaqt_counts_lr
  positional_bias = NULL,     # "3p", "5p", or NULL
  n_bins = 50L,
  prior_model = NULL,
  do_umi_correction = TRUE,   # Default for single-cell UMI data
  max_iter = 100L,
  tolerance = 1e-4,
  prior_start = 25L,
  convergence_start = 25L,
  verbose = TRUE
)
```

### Example: Quantify All Clusters

```r
# Quantify all clusters
results <- mpaqt_quant_sc(
  index = index,
  sr_counts_list = sr_list,
  lr_counts_list = lr_list,        # Optional
  positional_bias = "3p",          # Shared bias correction
  do_umi_correction = TRUE,        # Default; can be omitted
  max_iter = 100,
  verbose = TRUE
)

# Returns named list of mpaqt_quant_result objects
names(results)
# [1] "1" "2" "3" "4" "5" "all_clusters"

# Access individual cluster results
cluster1_tpm <- tpm(results[["1"]])
```

### Shared vs Per-Cluster Processing

Bias correction (`positional_bias`) uses the combined `"all_clusters"` data for more robust weight estimation. The weights are then applied to each cluster's EM algorithm. UMI correction is applied inside each `mpaqt_postquant()` call by default for single-cell workflows.

```r
# Under the hood, mpaqt_quant_sc does:

# 1. Pre-quantification on combined data
prequant <- mpaqt_prequant(
  index = index,
  sr_counts = sr_list[["all_clusters"]],
  positional_bias = "3p"
)

# 2. Apply to each cluster
for (cluster in names(sr_list)) {
  result <- mpaqt_postquant(
    index = index,
    sr_counts = sr_list[[cluster]],
    prequant = prequant,          # Shared weights
    do_umi_correction = TRUE      # Default for mpaqt_quant_sc()
  )
}
```

---

## Stage 5: Saving Results

### Save All Cluster Results

```r
# Save individual cluster results + combined matrix
mpaqt_save_result_sc(
  results = results,
  output_dir = "results/",
  prefix = "sample"
)

# Creates:
#   results/sample.1.quant.rds
#   results/sample.1.quant.csv
#   results/sample.2.quant.rds
#   results/sample.2.quant.csv
#   ...
#   results/sample.combined.csv  # TPM matrix
```

### Extract Combined TPM Matrix

```r
# Create TPM matrix: transcripts x clusters
combined_tpm <- combine_quant_results(results)

head(combined_tpm)
#            transcript_id        1        2        3
# 1: ENST00000456328.2 0.123456 0.234567 0.345678
# 2: ENST00000450305.2 2.345678 1.234567 3.456789
# 3: ENST00000488147.1 0.000000 0.012345 0.000000
```

---

## Complete Example Workflow

### Using CLI

```bash
#!/bin/bash
# Complete single-cell MPAQT workflow

# Stage 1: Create index (run once per reference)
mpaqt index \
  --transcriptome reference/gencode.v44.transcripts.fa \
  --annotation reference/gencode.v44.annotation.gtf \
  --output reference/mpaqt.index.rds \
  --read-length 91 \
  --threads 8

# Stage 2: Process single-cell short reads
mpaqt prepare-sr reference/mpaqt.index.rds \
  --fastq-r1 data/sample_R1.fastq.gz \
  --fastq-r2 data/sample_R2.fastq.gz \
  --technology 10xv3 \
  --output-dir results/ \
  --threads 8

# Stage 3: Quantify by cluster
mpaqt quant-sc reference/mpaqt.index.rds results/mpaqt.sr.rds \
  --clusters analysis/clusters.csv \
  --bias 3p \
  --output-dir results/

# Output files:
#   results/cluster_*.quant.rds   - Per-cluster results
#   results/combined.quant.csv    - TPM matrix (transcripts x clusters)
```

### Using R API

```r
library(mpaqt)

#===============================================================================
# STAGE 1: Index (create once, reuse)
#===============================================================================

index <- mpaqt_index(
  annotation = "reference/gencode.v44.annotation.gtf",
  transcriptome = "reference/gencode.v44.transcripts.fa",
  output_file = "reference/mpaqt.index.rds",
  read_length = 91,
  threads = 8
)

#===============================================================================
# STAGE 2: Process Single-Cell Short Reads
#===============================================================================

# Load cluster assignments from Seurat/scanpy analysis
# clusters.csv: barcode,cluster

sr_list <- mpaqt_prepare_short_reads_sc(
  index = index,
  fastq_1 = "data/sample_R1.fastq.gz",
  fastq_2 = "data/sample_R2.fastq.gz",
  clusters_file = "analysis/clusters.csv",
  technology = "10xv3",
  output_dir = "results/",
  threads = 8
)

#===============================================================================
# STAGE 3: Process Long Reads (Optional)
#===============================================================================

# If you have single-cell long-read data (e.g., MAS-ISO-seq)
lr_list <- mpaqt_prepare_long_reads_sc(
  index = index,
  count_file = "data/lr_counts.csv",
  clusters_file = "analysis/clusters.csv",
  output_dir = "results/"
)

#===============================================================================
# STAGE 4: Quantify All Clusters
#===============================================================================

results <- mpaqt_quant_sc(
  index = index,
  sr_counts_list = sr_list,
  lr_counts_list = lr_list,
  positional_bias = "3p",
  verbose = TRUE
)

#===============================================================================
# STAGE 5: Extract and Save Results
#===============================================================================

# Save everything
mpaqt_save_result_sc(results, "results/", prefix = "sample")

# Get combined TPM matrix for downstream analysis
tpm_matrix <- combine_quant_results(results)

# Export for use in Seurat/scanpy
data.table::fwrite(tpm_matrix, "results/transcript_tpm_by_cluster.csv")
```
